Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2018-05-24 RNARedPrint-Bioinfo2018.tex
Minor edits of 4.2 and suggested fix of MFE vs. G in MultiDefect function

\PassOptionsToPackage{utf8}{inputenc}
\documentclass{bioinfo}
\copyrightyear{2015} \pubyear{2015}

@@

\usepackage{todonotes}

@@

\newcommand{\TargetE}{E^{\star}}
\newcommand{\MFE}{\text{\rm MFE}}

@@

@@

%To reduce the cost of the precomputation, we used a stacking energy model
%as a proxy for the Turner model. Indeed, we observed a strong linear
%dependency between the Turner energy and the
%emphasize that only the simple stacking energies are directly targeted
%(at $\pm$10\% tolerance, except for a few hard instances). Due to our
%linear adjustment, we finally achieve well-controlled energies in the
%realistic Turner energy model.
\begin{figure*}[t]
  \begin{center}
    %{\sf \bfseries A}\
    \includegraphicstop[width=0.99\textwidth]{Figs/energy_shift.pdf}\hfill
    %{\sf \bfseries B}\ \includegraphicstop[width=0.55\textwidth]{Figs/PKB00211_PKB00239_0_energy_distribution}
  \end{center}
  \caption{%
    Targeting specific energies for pseudoknotted structures using
    multi-dimensional Boltzmann sampling.  (A) Linear fits between the
    energies in the stacking model to the realistic pseudoknot energy
    model by Dirks and Pierce (D\&P) for sampled sequences; the good
    match enables more efficient targeting of Turner energies based on
    targeting stacking model energies.  We show the fits for both
    target structures S1 and S2 of subfigure B.  (B) D\&P energy
    distributions for two target structures S1 and S2. We show the
    results of targeting the respective free energies $(-30,-20)$,
    $(-30,-30)$, $(-25,-25)$, and $(-20,-30)$ for the target
    structures (annotated in red, yellow, and blue
    resepectively)---this demonstrates the effectivity of our adaptive
    multi-dimensional Boltzmann sampling procedure; moreover, for
    comparison, the distributions for uniform and Boltzmann
    samples---respectively associated with homogenous weights $1$ and
    $e^\beta$.
%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
  }
  \label{fig:energydist-pk}
\end{figure*}
\subsection{Generating high-quality seeds for further optimization}
Next, we empirically evaluated the capacity of \ourprog{} to generate seed sequences for multiple (pseudoknotted) target structures, possibly followed by subsequent local optimizations. As a baseline for comparison, we considered \RNAblueprint~\citep{Hammer2017}, the current leading tool for multiple design. 
As a quality measure, we used the multi-stable design objective function introduced by~\citet{Hammer2017}, defined as:
% \begin{align}
%   f(S)= \quad & \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - G(S))\notag\\
%    & +\ \xi \frac{2}{k(k-1)} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
% \end{align}
\begin{align}
  \label{eq:blueprintobjective}
    \Obj(S)= \quad & \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - \MFE(S))\notag\\
   & +\frac{1}{2\binom{k}{2}} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
\end{align}
where $\MFE(S)$ is the Minimum Free-Energy of $S$, computed by {\tt RNAfold}~\citep{Hofacker1994} or approximated by {\tt HotKnots}~\citep{Ren2005} for pseudoknotted targets.
Intuitively, the first term of \Obj{} captures how distant from the $\MFE$ the targets are, while the second term penalizes the dispersion of targets. \Obj{} is best (minimized) when all targets simultaneously achieve the $\MFE$ of the sequence.

We considered a benchmark of six sets of target structures described in \citet{Taneda2015}: {\sf 2str}, {\sf 3str}, and {\sf 4str} consist of non-pseudoknotted structures, while {\sf PK60}, {\sf PK80}, and {\sf LE80} contain pseudoknotted structures.
%%For each instance of the \Software{Modena}
%benchmark~\citep{Taneda2015}, we generated $1\,000$ seed sequences at these energies.
%Our procedure is tailored to produce sequences with similar Turner
%energy that favor the stability of the target structures having
%moderate \GCb\%; all these properties are desirable objectives
%for RNA design. 
Using \ourprog{}, we generated at least 1\,000 seeds per instance for each instance 
of the benchmark. We used target energies that were computed by averaging the energy over samples at
higher weights $e^{\beta}$, and set the targeted \GCb\%
to 60\%.
 We compared the \Obj{} value of 
these sequences against that of seed sequences, uniformly sampled 
using \RNAblueprint{}.
Moreover, for both sets, we used an adaptive greedy
walk~\citep{Hammer2017} to \Def{optimize} the \Obj{} function. At
each step, the local search resamples (uniformly at random) the positions 
of a randomly selected component in the base pair dependency graph, 
accepting the modification only if it results in a gain. The greedy descent
was run for 500 steps in the case of pseudoknot-free datasets {\sf 2str}, {\sf 3str}, and
{\sf 4str}; and 200 steps for the pseudoknotted ones {\sf PK60}, {\sf PK80}, and {\sf LE80}.

The results, shown in Fig.~\ref{fig:benchmark-results}, confirm 
that Boltzmann-sampled sequences outperform uniform seeds on every dataset, 
leading to average improvements in \Obj{} values ranging from $7.26$ 
({\sf LE80}) to $16.05$  ({\sf 2str}) units. Remarkably, this improvement is
be observed for both of the terms in \Obj{} (see Supp. Mat.~\ref{sec:improvements}). 
This means that \ourprog{} produces sequences whose targets are both substantially closer to their MFE, and more 
similarly stable across targets. 
In fact, for every sequence in our benchmark, consisting of  332 sets of 
target structures in total, we observed better \Obj{} for Boltzmann-weighted sampling 
than for uniform sampling (see Supp. Mat.~\ref{sec:validity}). 
Notably, \ourprog{} remains 
unfazed by the presence of pseudoknots, the main problem being the computation 
of the \Obj{} function, which requires a costly free-energy minimization 
in the presence of pseudoknots~\citep{Sheikh2012}.

>>>>

\PassOptionsToPackage{utf8}{inputenc}
\documentclass{bioinfo}
\copyrightyear{2018} \pubyear{2018}

@@

%\usepackage{todonotes}

@@

\newcommand{\TargetE}{E^{\star}}
%\newcommand{\MFE}{\text{\rm MFE}}
\newcommand{\EnsE}{\text{\rm G}} %ensemble energy

@@

%% fix some weirdness of the bioinfo class (footer overlapping text on page 1)
\makeatletter
\setlength\footskip{30\p@}          % space above footer line
\makeatother

@@

%To reduce the cost of the precomputation, we used a stacking energy model
%as a proxy for the Turner model. Indeed, we observed a strong linear
%dependency between the Turner energy and the
%emphasize that only the simple stacking energies are directly targeted
%(at $\pm$10\% tolerance, except for a few hard instances). Due to our
%linear adjustment, we finally achieve well-controlled energies in the
%realistic Turner energy model.
\begin{figure*}[t]
  \begin{center}
    %{\sf \bfseries A}\
    \includegraphicstop[width=0.99\textwidth]{Figs/energy_shift.pdf}\hfill
    %{\sf \bfseries B}\ \includegraphicstop[width=0.55\textwidth]{Figs/PKB00211_PKB00239_0_energy_distribution}
  \end{center}
  \caption{%
    Targeting specific energies for pseudoknotted structures using
    multi-dimensional Boltzmann sampling.  A) Linear fits between the
    energies in the stacking model to the realistic pseudoknot energy
    model by Dirks and Pierce (D\&P) for sampled sequences; the good
    match enables more efficient targeting of Turner energies based on
    targeting stacking model energies.  We show the fits for both
    target structures S1 and S2 of subfigure B.  B) D\&P energy
    distributions for two target structures S1 and S2. We show the
    results of targeting the respective free energies $(-30,-20)$,
    $(-30,-30)$, $(-25,-25)$, and $(-20,-30)$ for the target
    structures (annotated in red, yellow, and blue
    resepectively)---this demonstrates the effectivity of our adaptive
    multi-dimensional Boltzmann sampling procedure; moreover, for
    comparison, the distributions for uniform and Boltzmann
    samples---respectively associated with homogenous weights $1$ and
    $e^\beta$.
%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
  }
  \label{fig:energydist-pk}
\end{figure*}
\subsection{Generating high-quality seeds for further optimization}
Next, we empirically evaluated the capacity of \ourprog{} to generate seed sequences for multiple (pseudoknotted) target structures, possibly followed by subsequent local optimizations. As a baseline for comparison, we considered \RNAblueprint~\citep{Hammer2017}, the current leading tool for multiple design. 
As a quality measure, we used the multi-stable design objective function introduced by~\citet{Hammer2017}, defined as:
% \begin{align}
%   f(S)= \quad & \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - G(S))\notag\\
%    & +\ \xi \frac{2}{k(k-1)} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
% \end{align}
\begin{align}
  \label{eq:blueprintobjective}
    \Obj(S)= \quad & \frac{1}{k} \sum_{\ell=1}^{k} (E(S, R_\ell) - \EnsE(S))\notag\\
   & +\frac{1}{2\binom{k}{2}} \sum\limits_{1\leq\ell<j\leq k}|E(S,R_\ell) - E(S,R_j)|.
\end{align}
where $\EnsE(S)$, the \emph{ensemble free energy} of $S$, is computed
by {\tt RNAfold}~\citep{Hofacker1994} in the pseudoknot-free case; for
pseudoknotted targets, $\EnsE(S)$ is approximated by the \emph{minimum
  free energy} of S as estimated by {\tt HotKnots} in the energy model
of~\citet{dirks-pierce-03}.  Intuitively, the first term of \Obj{}
captures the distance of the targets from the ensemble free energy,
while the second term penalizes the dispersion of targets; \Obj{} is
best (minimal) when all targets simultaneously achieve the minimum
free energy of the sequence.

We considered a benchmark of six sets of target structures described in \citet{Taneda2015}: {\sf 2str}, {\sf 3str}, and {\sf 4str} consist of non-pseudoknotted structures, while {\sf PK60}, {\sf PK80}, and {\sf LE80} contain pseudoknotted structures.
%%For each instance of the \Software{Modena}
%benchmark~\citep{Taneda2015}, we generated $1\,000$ seed sequences at these energies.
%Our procedure is tailored to produce sequences with similar Turner
%energy that favor the stability of the target structures having
%moderate \GCb\%; all these properties are desirable objectives
%for RNA design. 
Using \ourprog{}, we generated at least 1\,000 seeds per instance for
each instance of the benchmark. We used target energies that were
computed by averaging the energy over samples at higher weights
$e^{\beta}$, and set the targeted \GCb\% to 60\%.  We compared the
\Obj{} value of these sequences against that of seed sequences,
uniformly sampled using \RNAblueprint{}.  Moreover, for both sets, we
used an adaptive greedy walk~\citep{Hammer2017} to \Def{minimize} the
\Obj{} function. At each step, the local search resamples (uniformly
at random) the positions of a randomly selected component in the base
pair dependency graph, accepting the modification only if it results
in a gain. We performed 500 greedy descent steps in the case of
pseudoknot-free datasets {\sf 2str}, {\sf 3str}, and {\sf 4str}; and
200 steps for the pseudoknotted ones {\sf PK60}, {\sf PK80}, and {\sf
  LE80}.

The results, shown in Fig.~\ref{fig:benchmark-results}, confirm that
Boltzmann-sampled sequences outperform uniform seeds on every dataset,
leading to average improvements in \Obj{} values ranging from $7.26$
({\sf LE80}) to $16.05$ ({\sf 2str}) units. Remarkably, this
improvement is be observed for both of the terms in \Obj{} (see
Supp. Mat.~\ref{sec:improvements}).  This means that \ourprog{}
produces sequences whose targets are both substantially closer to
their MFE, and more similarly stable across targets.  In fact, for
every sequence in our benchmark, consisting of 332 sets of target
structures, we observed better \Obj{} for Boltzmann sampling than for
uniform sampling (see Supp. Mat.~\ref{sec:validity}).  Notably,
\ourprog{} performs equally well in the presence of pseudoknots;
the difficulty rather lies in the computation of the \Obj{} function, since
free-energy minimization is costly in the presence of
pseudoknots~\citep{Sheikh2012} and good implementations are scarce.
